The X-linked intellectual disability gene KDM5C is a sex-biased brake against germline programs in somatic lineages

### 2024-11-05 submission/manuscript.Rmd
abstract edits

The division of labor among cellular lineages is a pivotal step in the evolution of multicellularity. In mammals, the soma-germline boundary is formed during early embryogenesis, when genes that establish germline identity are repressed in somatic lineages through DNA and histone modifications at promoter CpG islands (CGI). Somatic misexpression of germline genes is a signature of cancer and observed in select neurodevelopmental disorders. However, it is currently unclear if all germline genes use the same repressive mechanisms and if factors like development and sex influence their dysregulation. Here, we examine how cellular context influences the formation of somatic tissue identity in mice lacking lysine demethylase 5c (KDM5C), an X chromosome eraser of histone 3 lysine 4 di and tri-methylation (H3K4me2/3). We found KDM5C is a crucial regulator of tissue identity, as male _Kdm5c_ knockout (-KO) mice aberrantly express many liver, muscle, and germline genes within the brain. By developing a comprehensive list of mouse germline-enriched genes, we found the _Kdm5c_-KO brain aberrantly expresses late-stage spermatogenesis genes while _Kdm5c_-KO epiblast-like cells (EpiLCs) primarily expressed key regulators of germline fate, such as _Dazl_ and _Stra8_. KDM5C-mediated germline gene repression is sexually dimorphic, as female EpiLCs require a higher dose of KDM5C to maintain germline silencing. KDM5C represses germline genes during ESC to EpiLC differentiation and is recruited to a subset of CGI-containing germline gene promoters to facilitate DNA CpG methylation. However, the majority of mouse germline gene promoters did not harbor CGIs and were unbound by KDM5C, including late-stage spermatogenesis genes that are expressed in _Kdm5c_-KO cells through ectopic activation by RFX transcription factors. Altogether, these data demonstrate KDM5C's fundamental role in tissue identity and indicate that KDM5C acts as a sex-biased brake against runaway activation of germline developmental programs in somatic lineages.

>>>>

The division of labor among cellular lineages is a pivotal step in the evolution of multicellularity. In mammals, the soma-germline boundary is formed during early embryogenesis, when genes that establish germline identity are repressed in somatic lineages through DNA and histone modifications at promoter CpG islands (CGI). Somatic misexpression of germline genes is a signature of cancer and observed in select neurodevelopmental disorders. However, it is currently unclear if all germline genes use the same repressive mechanisms and if factors like development and sex influence their dysregulation. Here, we examine how cellular context influences the formation of somatic tissue identity in mice lacking lysine demethylase 5c (KDM5C), an X chromosome eraser of histone 3 lysine 4 di and tri-methylation (H3K4me2/3). We found male _Kdm5c_ knockout (-KO) mice aberrantly express many tissue-specific genes within the brain, the majority of which are unique to the germline. By developing a comprehensive list of mouse germline-enriched genes, we observed _Kdm5c_-KO cells aberrantly express key drivers of germline fate during early embryogenesis but late-stage spermatogenesis genes within the mature _Kdm5c_-KO brain. KDM5C binds germline gene promoter CGIs to facilitate DNA CpG methylation as embryonic stem cells differentiate into epiblast-like cells (EpiLCs). However, the majority of late-stage spermatogenesis genes expressed within the _Kdm5c_-KO brain did not harbor promoter CGIs. These CGI-free germline genes were not bound by KDM5C and instead expressed through ectopic activation by RFX transcription factors. Furthermore, KDM5C-mediated germline gene repression is sexually dimorphic, as female EpiLCs require a higher dose of KDM5C to maintain germline silencing.

Altogether, these data revealed distinct regulatory classes of germline genes and sex-biased silencing mechanisms in somatic cells.
